Template for indicator documentation

### 2023-09-12 DRAFT_NDVI_wetland.Rmd
fsøgjs
ndvi_wetland_continued

# Functional plant indicators, wetland ecosystems (Moisture, Light, pH, Nitrogen) {#Functional-plant-indicators-wetland}

@@

#### NDVI data
```{r}
## Sentinel-2
# join nin.wetland & Sentinel NDVI data
SentinelNDVI.wetland <- full_join(nin.wetland, df.s, by="id")
#summary(SentinelNDVI.wetland)
SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
  mutate(hovedoekosystem = as.factor(hovedoekosystem),
         hovedtype = as.factor(hovedtype),
         ninkartleggingsenheter = as.factor(ninkartleggingsenheter), 
         lokalitetskvalitet = as.factor(lokalitetskvalitet),
         tilstand = as.factor(tilstand),
         area_meters = st_area(SentinelNDVI.wetland))
#summary(SentinelNDVI.wetland)
# get rid of NAs (i.e. NDVI cells that were not in wetland polygons)
SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(!is.na(hovedtype))
SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(!is.na(mean))
#summary(SentinelNDVI.wetland)
# get rid of any nin-polygons smaller than the Sentinel grid cell size (100 sqm)
dim(SentinelNDVI.wetland)
SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(as.numeric(area_meters) >= 100)
dim(SentinelNDVI.wetland)
# split date into year, month & day
SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
  dplyr::mutate(year = lubridate::year(date), 
                month = lubridate::month(date), 
                day = lubridate::day(date))
# add column for sub-ecosystem types
SentinelNDVI.wetland <- SentinelNDVI.wetland %>% mutate(subtype = substring(ninkartleggingsenheter, 4),
                                                        subtype = str_remove(subtype, '-'))
# we are using max NDVI per year in every NiN polygon
SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm=TRUE))

@@

```{r}
## how are polygon sizes distributed?
summary(SentinelNDVI.wetland$area_meters)

hist(SentinelNDVI.wetland$area_meters,xlim=c(0,20000),breaks=50000)
abline(v=100,lty=2)

## how does NDVI vary over the years (all years)
SentinelNDVI.wetland %>%
  ggplot( aes(x=year, y=mean )) + 
  geom_point() +
  facet_grid( tilstand~hovedtype)
# 2022 stands out with the highest NDVI values missing
# the pattern is strange, suggest to omit 2022 for this analysis

SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(year != '2022')

## NDVI across hovedtyper (only for NDVI years data matching NiN-mapping years)
SentinelNDVI.wetland %>%
  filter(year == kartleggingsaar) %>%
  ggplot( aes(x=tilstand, y=mean )) + 
  geom_violin() +
  facet_wrap( ~hovedtype)
# NDVI largely varies between hovedtyper: higher in seminat wet-meadows (V10), lowest in ombrotrophic bogs (V3)
# @Tilstand: in seminat types, kaldkilde, and bogs ndvi seems to increase as condition deteriorates, in fens not much of a pattern

## looking at subtypes for polygons in good condition only (and only for NDVI data matching NiN-mapping)
SentinelNDVI.wetland %>%
  filter(tilstand == 'God') %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x=subtype, y=mean )) + 
  geom_violin() +
  facet_wrap( ~hovedtype)
# NDVI varies between ecosystem subtypes
# -at least equally much as between main ecosystem types
# -stronger than between condition classes in previous plots

## looking at ndvi in different condition classes, separately for main types and regions  (only for NDVI data matching NiN-mapping)
SentinelNDVI.wetland %>%
  group_by(id, year) %>%

  filter(year == kartleggingsaar) %>%
  ggplot( aes(x=tilstand, y=mean )) + 
  geom_violin() +
  facet_grid( region~hovedtype)
# V4 has lacking data for quite some condition classes in most regions

@@

???

@@

# concatenate the three Satellite objects
allSatNDVI.wetland <- rbind(
SentinelNDVI.wetland[,!names(SentinelNDVI.wetland) %in% "subtype"],
ModisNDVI.wetland,
LandsatNDVI.wetland[,!names(LandsatNDVI.wetland) %in% "column_label"]
)
# add variable for Satellite indentity
allSatNDVI.wetland$Sat <- c(
  rep("Sentinel",nrow(SentinelNDVI.wetland)),
  rep("Modis",nrow(ModisNDVI.wetland)),
  rep("Landsat",nrow(LandsatNDVI.wetland))
)
allSatNDVI.wetland$Sat <- factor(allSatNDVI.wetland$Sat,levels=c("Sentinel","Modis","Landsat"))
levels(allSatNDVI.wetland$Sat)
```
Now we can plot the time series for each main ecosystem type, and for each satellite separately  
```{r}
# plot 
allSatNDVI.wetland %>%
  ggplot( aes(x=year_jit, y=mean, color=Sat )) + 
  geom_point() +
  facet_wrap(~hovedtype, ncol = 1)
```
It is quite obvious that the NDVI values from the three satellites are not quantitatively comparable. They vary both in their placement along the y-axis and in their variance. Modis data are absent from the V4-ecosystem type because of a too large pixel size compared to the NiN-polygon sizes. In V10, Modis is poorly represented for the same reason.

>>>>

# NDVI (greenness) indicator, wetland ecosystems  {#NDVI-indicator-wetland}

@@

#### NDVI data
```{r}
## Sentinel-2
# join nin.wetland & Sentinel NDVI data
SentinelNDVI.wetland <- full_join(nin.wetland, df.s, by="id")
#summary(SentinelNDVI.wetland)
SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
  mutate(hovedoekosystem = as.factor(hovedoekosystem),
         hovedtype = as.factor(hovedtype),
         ninkartleggingsenheter = as.factor(ninkartleggingsenheter), 
         lokalitetskvalitet = as.factor(lokalitetskvalitet),
         tilstand = as.factor(tilstand),
         area_meters = st_area(SentinelNDVI.wetland))
#summary(SentinelNDVI.wetland)
# get rid of NAs (i.e. NDVI cells that were not in wetland polygons)
SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(!is.na(hovedtype))
SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(!is.na(mean))
#summary(SentinelNDVI.wetland)
# get rid of any nin-polygons smaller than the Sentinel grid cell size (100 sqm)
dim(SentinelNDVI.wetland)
SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(as.numeric(area_meters) >= 100)
dim(SentinelNDVI.wetland)
# split date into year, month & day
SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
  dplyr::mutate(year = lubridate::year(date), 
                month = lubridate::month(date), 
                day = lubridate::day(date))
# add column for sub-ecosystem types
SentinelNDVI.wetland <- SentinelNDVI.wetland %>% mutate(subtype = substring(ninkartleggingsenheter, 4),
                                                        subtype = str_remove(subtype, '-'),
                                                        subtype = str_remove(subtype, '-'))
# we are using max NDVI per year in every NiN polygon
SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm=TRUE))

@@

```{r}
## how are polygon sizes distributed?
summary(SentinelNDVI.wetland$area_meters)
hist(SentinelNDVI.wetland$area_meters,xlim=c(0,20000),breaks=50000)
abline(v=100,lty=2)
# wetland polygons stretch a large number of sizes from just above 100 sqm to mor ethan 4 million sqm. 75% of all polygons are below 14264 sqm.

## how does NDVI vary over the years (all years)
SentinelNDVI.wetland %>%
  ggplot( aes(x=year, y=mean )) + 
  geom_point() +
  facet_grid( tilstand~hovedtype)
# 2022 stands out with the highest NDVI values missing
# the pattern is strange, suggest to omit 2022 for this analysis
SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(year != '2022')

## looking at major types for polygons in good condition only (and only for NDVI data matching NiN-mapping)
SentinelNDVI.wetland %>%
  filter(tilstand == 'God') %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x=hovedtype, y=mean )) + 
  geom_violin() +
  geom_point(size=0.7, shape=16, color="grey")
# NDVI varies between hovedtyper: highest in seminat wet-meadows (V10), lowest in ombrotrophic bogs (V3)

## looking at subtypes for polygons in good condition only (and only for NDVI data matching NiN-mapping)
SentinelNDVI.wetland %>%
  filter(tilstand == 'God') %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x=subtype, y=mean )) + 
  geom_violin() +
  geom_point(size=0.7, shape=16, color="grey") +
  facet_wrap( ~hovedtype)
# NDVI varies between ecosystem subtypes in V1. For the other types things seem more constant, but there's also less data and less minor ecosystem types present in the data.

## NDVI across regions for polygons in good condition only (for major types and NDVI years data matching NiN-mapping years)
SentinelNDVI.wetland %>%
  filter(tilstand == 'God') %>%
  filter(year == kartleggingsaar) %>%
  filter(!is.na(region)) %>%
  ggplot(aes(x=region, y=mean )) + 
  geom_violin() +
  geom_point(size=0.7, shape=16, color="grey") +
  facet_wrap( ~hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust=0.2))
# no obvious trends in geography

## NDVI across condition classes (for major types and NDVI years data matching NiN-mapping years)
SentinelNDVI.wetland %>%
  filter(year == kartleggingsaar) %>%
  ggplot( aes(x=tilstand, y=mean )) + 
  geom_violin() +
  geom_point(size=0.7, shape=16, color="grey") +
  facet_wrap( ~hovedtype)
# in seminat types (V9,V10), kaldkilde (V4), and bogs (V3) ndvi seems to increase as condition deteriorates, in fens (V1) not much of a pattern

@@

In semi-natural wetland ecosystems (V9, V10) NDVI increases as ecosystem condition deteriorates, and this is consistent for the entire country.
In natural fens (V1) NDVI increases in the South but decreases in the North as ecosystem condition deteriorates. In bogs (V3) there is no consistent pattern.

@@

# concatenate the three Satellite objects
allSatNDVI.wetland <- rbind(
SentinelNDVI.wetland[,!names(SentinelNDVI.wetland) %in% "subtype"],
ModisNDVI.wetland,
LandsatNDVI.wetland[,!names(LandsatNDVI.wetland) %in% "column_label"]
)
# add variable for Satellite identity
allSatNDVI.wetland$Sat <- c(
  rep("Sentinel",nrow(SentinelNDVI.wetland)),
  rep("Modis",nrow(ModisNDVI.wetland)),
  rep("Landsat",nrow(LandsatNDVI.wetland))
)
allSatNDVI.wetland$Sat <- factor(allSatNDVI.wetland$Sat,levels=c("Sentinel","Modis","Landsat"))
levels(allSatNDVI.wetland$Sat)
```
Now we can plot the time series for each main ecosystem type, and for each satellite separately  
```{r}
# plot 
allSatNDVI.wetland %>%
  ggplot( aes(x=year_jit, y=mean, color=Sat )) + 
  geom_point() +
  facet_wrap(~hovedtype, ncol = 1)
```
It is quite obvious that the NDVI values from the three satellites are not quantitatively comparable. They vary both in their placement along the y-axis and in their variance. Modis data are absent from the V4-ecosystem type because of a too large pixel size compared to the NiN-polygon sizes. In V10, Modis is poorly represented for the same reason.
